Seasonality in dominant upper-ocean dynamics at submesoscales

### 2016-05-09 writeup/rocha_etal.tex
Adds significant text on sections 4 and 5

\section{The LLC2160 model output}
We use the output of the latitude-longitude polar cap (LLC)
MITgcm  numerical simulation. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and high-resolution surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years form March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. The model output was provided by D. Menemenlis
(JPL/NASA).

@@

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows project are supra-inertial and project
on scales smaller than 100 km: daily-averaging the velocity fields or low-pass
filtering (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with that
of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.

@@

@@

\section{Projection on time scales}
To better understand the dynamics the projection of these flows on horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance.

>>>>

\section{The LLC2160 model output}
We use the output of the latitude-longitude polar cap (LLC)
MITgcm  numerical simulation. The output analyzed here, the LLC2160,
is the 1/24$^\circ$ --- 90-vertical-levels simulation from a hierarchy of global
forward numerical solutions on a LLC grid \citep{forget_etal2015} that were
spun up from a ECCO2 adjoint-method state estimate. The LLC2160 was forced by
tides and high-resolution surface fluxes similarly to the LLC4320 simulation
used by \citet{rocha_etal2016} --- see their appendix D. We chose the second-higher
resolution  simulation (LLC2160) of the LLC hierarchy because it was run for longer:
the LLC2160 output spans two years form March 2011 to April 2013. The model time-step
was 45s and the model variables were saved hourly. LLC hierarchy setup
and control files are available online
(at \texttt{http://mitgcm.org/viewvc/MITgcm/MITgcm$\_$contrib/
llc$\_$hires}). For the LLC hierarchy, the sea-surface height is calculated
using a linearized equation. While the nominal horizontal resolution is 1/24$^\circ$,
 horizontal wavenumber spectra suggests an effective resolution of about 20 km.

@@

@@

The horizontal divergent part of the flow is significant: the RMS is $\delta/f > 0.1$).
Most of these horizontal divergent flows project are supra-inertial and project
on scales smaller than 100 km: daily-averaging the velocity fields or low-pass
filtering (100 km cutoff) reduces the horizontal divergence by about 70$\%$.
Furthermore, the seasonal cycle in horizontal divergence is out-of-phase with that
of vertical vorticity: the horizontal divergence is about 30$\%$ larger in late
Summer/early Fall than it is in late Winter/early Spring. The divergence associated
with submesoscale subinertial eddies is small and its seasonal cycle is in phase
with that of vorticity (see green lines).  The seasonal cycle of the horizontal
divergence is roughly in phase with the seasonal cycle of the near-surface (0-50m)
stratification.
\begin{figure*}[ht]
\begin{center}
\includegraphics[width=1.\textwidth]{figs/fig1.pdf}
 \caption{Snapshots of surface vorticity (a and b) and a transect
          of potential density at 165$^\circ$E (c and d). The snapshots were
          taken at 00:00 UTC.}
 \label{fig1}
 \end{center}
 \end{figure*}

@@

To better understand the dynamics the projection of these flows on horizontal
scales, we calculate wavenumber spectra of kinetic energy and sea-surface height
(SSH) variance. For simplicity we calculate zonal wavenumber spectra at various
latitudes. The two-year time averaged and any spatial linear trend were
removed before calculating discrete Fourier transforms.Statistical significance
is obtained by averaging over latitude and time. Using conservative temporal and
spatial decorrelation estimate of 10 days and 250 km, we obtain roughly 50
independent realizations of the spectra.

Figure \ref{fig3}a depicts the horizontal wavenumber spectra of surface KE in
April and October. At scales larger than 20km, the spectra based on hourly
velocity snapshots are nearly indistinguishable within 95$\%$ confidence level
(compare solid lines in figure \ref{fig3}a). Consistent with the results of
\citet{rocha_etal2016} who analyzed the LLC4320 output in Drake Passage,
there is significant high-frequency variabilty at submesoscales. Daily-averaging
the velocity field suppresses spatial variability at scales small than about 250
km both in April and October (compare solid lines against dashed lines in figure
\ref{fig3}a). But this suppression is more dramatic in October, when the horizontally
divergent flows peak. At scales
smaller than 100km, 40$\%$ of the surface KE in April a is accounted for by suprainertial
flows as opposed to 77$\%$ in October. The seasonal of subinertial submesoscale flows
is significant, consistent with the results of \citet{sasaki_etal2014}, which are
based on snapshots of a model without tidal forcing.

These high-frequency submesoscale flows significantly project on the sea-surface.
In October, when the horizontally divergent flows peak and the submesoscale
balanced flows are minimum, there is a dramatic difference between the spectra
based on hourly and daily-averaged SSH: at scales smaller than 100 km, the spectra
of hourly SSH roughly follows a $-2$ power-law, whereas the spectra of daily-averaged
SSH roughly follows a $-5$ power-law. At scales
smaller than 100km, 44$\%$ of the surface KE in April is accounted for by suprainertial
flows as opposed to 86$\%$ in October! The out-of-phase seasonal cycle of subinertial
and suprainertial flows conspire to yield no significant seasonality in the spectra
of hourly SSH.

@@

\section{Projection on time scales}
To better understand the which flows project onto high-wavenumbers, the projection
of these flows on time scales, we calculate frequency spectra of surface KE
(figure \ref{fig4}). To obtain statistically meaningful results, we average
spectral estimates along a section at 165$^\circ$E.  There is also about 50
independent realizations of spectral estimates.

The surface KE spectra of subinertial flows very roughly follow a $-2$ power-law.
The high-frequency flows are split near-inertial flows, tidal signals and their
higher harmonics. Hence these high-frequency flows are dominates by inertia-gravity
waves. (Note that the the inertial peak nearly merges with the
diurnal tide peak). There is statistically significant seasonality. The seasonal
cycle of the sub-inertial motions (at least for periods between 5-20 days)
is out-of-phase with the seasonal cycle of supra-inertial flows: the former
are more energetic in April and the latter in October.

Most of these high-frequency inertia-gravity waves project onto scales smaller
than 100 km. The surface KE spectra of 100-km-smoothed velocity field is significantly
suppressed in supra-inertial variability --- there are no statistically significant
changes at sub-inertial frequencies (see dashed lines in figure \ref{fig4}).

% say something about relative energy in smoother vs. unsmoothed fields.
% also say something about comparison with observations.
